Fire Safety Regulation Violations in Toronto Residential Properties: Analysis by Property Type and Inspection Duration

### 2024-12-04 paper/paper.qmd
Updated final completed version of paper

In recent years, fire safety inspections in residential buildings have gained increasing attention from the public due to their role in ensuring the safety of residents and reducing potential dangers caused by fire-related accidents. Toronto Fire Services (TFS) fire prevention and enforcement strategies, which are guided by the Fire Protection and Prevention Act, 1997 (FPPA) and regulations under it (such as Ontario Fire Code), aim to increase public safety and awareness [@fire_info]. In order to achieve this goal, TFS have been conducting inspections across a wide range of residential properties within the entire city to ensure properties' compliance with fire safety regulations. However, despite these efforts, there remains a limited understanding of the factors influencing the likelihood of fire safety violations across different property types and inspection durations. As a result, targeted interventions have not yet been effectively implemented for high-risk buildings. Therefore, this paper aims to analyze the impact of both inspection time interval and property type on the likelihood of violations in fire safety checks. By doing so, a model that provides predictions of violations based on specific features of the properties and inspection characteristics could be developed, improving TFS's ability to prioritize resources and focus on properties that require the most attention.

The focus of this study is to investigate how different property types and inspection time intervals contribute to fire safety violations in Toronto, and this paper uses Bayesian logistics model to conduct predictions. The predictor variables examined include the time interval of the inspection process, and the type of property being inspected. By exploring these factors, the paper aims to predict the likelihood of violations, conducting predictions of which property types are at higher risk for violations and how inspection duration might influence these outcomes. The estimand would be described in detail in @sec-estimand.

As included at the end of the analysis, the model predicts with high confidence that most properties would have a high probability of being observed to have a violation of fire safety regulations. Such prediction by the model is confident, highlighting the utility of the model in helping to conduct targeted inspections and interventions where they are most needed. Moreover, the analysis further highlights that different property types have varying impacts on the likelihood of fire safety violations. The model confidently predicts that among ten property types inspected, "High Rise," "Rooming House," and "Group Home" are more likely to have violations. As a result, these property types should receive more targeted inspections and interventions. Along with "Low Rise" and "Hotel & Motel," these property types show strong evidence of being associated with higher risks of violations. In contrast, the model indicates that "Nursing Home" has the lowest risk of violations among ten property types inspected and some uncertainty exist in the strength of this predicted relationship.

This research contributes to improving fire safety facilities in residential buildings in Toronto by providing empirical evidence and statistical predictions that supports the allocation of inspection resources based on both the property type and inspection time interval. By highlighting which property types with how long inspection duration are most at risk for fire safety violations, the findings offer a valuable approach to prioritizing inspections and interventions. The results of the study have broader implications for fire safety policy, as they could be used to improve future regulatory practices and resource allocation strategies. By ensuring that inspection resources are concentrated on properties that present the greatest risk, this study would enhance the effectiveness of fire safety efforts, and help to reduce potential fire accidents more efficiently.

@@

Appendix includes three parts:[Appendix -@sec-appx-survey] describes detailed idealized sampling methodology and survey designed for residents to collect fire safety feedback; [Appendix -@sec-appx-data] provides additional data details; [Appendix -@sec-model-details] presents additional details during modeling process, including posterior predictive check [Appendix -@sec-ppc], comparison between posterior and prior distributions [Appendix -@sec-pvp], and diagnostics [Appendix -@sec-diagnostics].

@@

The dataset captures data related to the results of fire safety inspections conducted by Toronto Fire Services (TFS) within highrise residential buildings, which is guided by the Fire Protection and Prevention Act, 1997 (FPPA) and regulations under the FPPA, like the Ontario Fire Code [@fire_info]. It indicates properties where violations have been found by an Inspector which are required to be fixed for compliance at the time with the Ontario Fire Code as well as properties with no observable violations [@fire_data].

The data collection process was conducted by TFS Fire Inspectors. Inspectors assess tangible aspects of fire safety based on observable and measurable attributes of the physical environment within the properties, such as proper storage of combustible materials, functionality of fire alarms and fire pumps, maintenance records for fire safety equipment and others in reference to the regulations under FPPA.

After the inspections were completed, the observations have been translated into records and written into collected data. The raw data entry presented by Open Data Toronto includes three main aspects: property identifier (such as address, property type), inspection information (such as inspection start and end dates), and inspection outcomes (pass or fail status for safety checks, and details of violations).

Note that this dataset only includes 'closed' cases, which are cases whose inspection processes have ended. The ongoing inspections were not available in the dataset, and therefore would lead to potential loss of information. This would be further evaluated in detail in the discussion section.

@@

The outcome variable for the model is the violation status of the properties. In other words, it is the variable that represents whether the properties inspected have been found violating fire safety regulations by an Inspector or not.

In the raw dataset, columns 'VIOLATION_FIRE_CODE', 'VIOLATION_DESCRIPTION' and 'VIOLATIONS_ITEM_NUMBER' indicates the outcome of the inspections. The previous two provides detailed information on fire code under which violation was noted. The latter one presents the order number of violations by code, and shows 0 if no violations observed.

Therefore, a new column 'violation' is mutated to make the analysis procedure easier. It contains numeric values of 1 and 0, which indicates violation observed and no violation observed respectively. The outcome variable is thus binary.

@fig-violation presents the visualized counts of properties that were found violations versus those that were not. The height of each bar shows the total number of cases corresponding to each category. We could see that the orange bar is significantly taller, indicating that among all cases selected for analysis, violations were reported more frequently in the dataset (9829 cases out of 13193). On the other hand, the blue bar is shorter, indicating there were fewer cases without violations (3364 cases out of 13193).

In conclusion, the data contains a higher number of cases with violations (1) compared to cases where no violations were observed (0), suggesting that violations are common.

```{r}
#| label: fig-violation
#| fig-cap: Counts of Properties that Were Found Violations versus Those that Were Not
#| echo: false
#| warning: false
#| message: false

# Create bar plot
cleaned_data |>
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  labs(
    x = "Violation Status",
    y = "Count",
    fill = "Violation Status" # Legend title
  ) +
  scale_fill_manual(
    values = c("skyblue", "orange"),
    labels = c("No Violation Observed (Denoted as 0)", "Violation Observed (Denoted as 1)")
    ) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

@@

### Inspection Time Interval

One predictor included in the model is the inspection time interval. It refers to the duration of the fire inspections conducted by the TFS Inspector for each case.

In the raw dataset, there are two columns recording the date: 'INSPECTIONS_OPENDATE' and 'INSPECTIONS_CLOSEDDATE'. Previous one is the date TFS opened the inspection file, whereas the latter one is the date TFS closed inspection file or date or date enforcement proceedings ended (whichever is later).

During the data cleaning process, the date is converted into time interval to enable easier analysis. A new column named 'date_num' is mutated by calculating the time difference between the end time and the start time of the inspection, in the unit of days.

@fig-date shows the distribution of the inspection time interval. X-axis represents the number of days of each inspection case, and Y-axis represents the counts. The distribution has peak at low intervals, and the highest bar is near 0 days, indicating that a large number of properties have very short inspection time intervals. The right-skewed inspection suggests that as the inspection interval increases, the corresponding counts decrease stably. The long tail of the right-hand side of the distribution means that there are properties with very long intervals (up to 1500 days or more).

```{r}
#| label: fig-date
#| fig-cap: Distribution of Inspection Time Interval
#| echo: false
#| warning: false
#| message: false

cleaned_data |> 
  ggplot(aes(x = date_num)) + 
  geom_histogram(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(x = "Inspection Time Interval (Days)",
       y = "Counts of Properties") +
  theme_minimal()
```

@@

The second predictor variable included in the model is the property type. It refers to the occupancy type of the property where the inspection took place, and is directly contains in the raw dataset.

To visualize the variable, two bar plots are created separately in @fig-type. The left panel presents the number of occurrences of "High Rise" and "Low Rise", with the bar representing "High Rise" relatively taller, indicating a higher count compared to low rise. This suggests that the inspection results and data collection focus more heavily on high-rise buildings.

The right panel includes the other categories of properties. Specifically, the types include: "Detention," "Group Home," "Group Home (VO)," "Hospital," "Hotel & Motel," "Nursing Home," "Residential Care," and "Rooming House." We could see that "Rooming House" has the highest count among these property types, whereas categories of "Detention" and "Group Home" have relatively lower counts, indicating that they are less represented in the dataset.

In general, there is a diverse distribution across the property types, with some being far less frequent than others. High-rise properties dominate overall, while rooming houses stand out within other categories. The dataset might reflect an emphasis on inspecting high-rise and rooming houses more than other property types.

```{r}
#| label: fig-type
#| fig-cap: Counts of Property Types
#| echo: false
#| warning: false
#| message: false
#| fig-subcap: ["High Rise versus Low Rise", "Other Types"]
#| layout-ncol: 2

@@

# Create bar plot for high rise properties versus low rise properties
data_rise |>
  ggplot(
    aes(x = as.factor(property_type), 
        fill = as.factor(property_type))
    ) +
  geom_bar(color = "black") +
  labs(
    x = "Property Type",
    y = "Count",
    fill = "Property Type"
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

# Create bar plot for other property types
data_other |>
  ggplot(
    aes(x = as.factor(property_type), 
        fill = as.factor(property_type))
    ) +
  geom_bar(color = "black") +
  labs(
    x = "Property Type",
    y = "Count",
    fill = "Property Type" 
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

@@

@fig-violationbytype depicts faceted bar plots showing the distribution of violation statuses (0 for no violation, 1 for violation) across different property types. Each subplot corresponds to a different property type.

In general, property types of "Group Home," "High Rise," "Low Rise," and "Rooming House," have a higher proportion of violations (1). The type "Hotel & Motel" has balanced counts between 0 and 1, but violations (1) are slightly more frequent. These property types may require more attention or stricter inspections.

The other property types, especially "Detention" and "Nursing Home", show a much lower proportion of violations, suggesting fewer relative violation ratio.

@@

cleaned_data |> 
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  facet_wrap(~ property_type, scales = "free_y") + # Create one subplot per property type
  labs(
    x = "Violation Status",
    y = "Count",
    fill = "Violation Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_blank(), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 8), 
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.spacing = unit(1, "lines")
  ) +
  scale_fill_manual(values = c("skyblue", "orange"))
```

@@

Firstly and most importantly, from the statistical standpoint, the goal is to estimate the the detection of violation from an inspection based on its time interval and the property type.

Secondly, from a more practical standpoint, by predicting which type of property with how long duration is more likely to be observed violations, more efficient inspection process could be constructed. Identifying certain property types that would more frequently violate the fire safety regulations could provide guidelines for interventions and help to establish targeted and focused actions towards them.

The model constructed in the paper is Bayesian logistics regression model designed to predict the violation status (denoted as violation) based on the fixed effect of inspection time interval (denoted as date_num) and property type (denoted as property_type). It is designed to predict the inspection results based on the two predictor variables.

Following sections define, explain and justify each model and the variables, as well as discuss underlying assumptions, potential limitations, software used to implement the model, and evidence of model validation and checking.

@@

The Bayesian model was implemented using the R programming language [@citeR], specifically utilizing the `rstanarm` package of @citerstanarm. This package provides an interface for fitting Bayesian regression models using Stan, and the models are fit using the package.

Define $y_i$ as the binary response variable for observation $i$. $y_i=1$ indicates that the case has been observed violations, while $y_i=0$ indicates that no violation occurs. It follows a Bernoulli distribution:

@@

Then define the inspection time interval for observation $i$ as date_num and the property type as property_type.

The Bayesian model could be defined by the following mathematical expressions:

@@

- $\text{logit}(p_i) = \log\left(\frac{p_i}{1 - p_i}\right)$ is the log-odds of $p_i$.
- $\beta_0$ is the intercept term of the Bayesian logistic regression, representing the baseline log-odds of a violation when 'date_num = 0' and 'property_type' is the reference category. Formula (2) specifies that $\beta_0$ follows normal distribution with mean equals 0 and variance equals $2.5^2$.
- $\beta_1$ is the coefficient for predictor variable 'date_num‘, representing the change in log-odds of a violation per unit increase in 'date_num', the inspection time interval. Formula (3) specifies that $\beta_1$ follows normal distribution with mean equals 0 and variance equals $2.5^2$.
- $\beta_2$ is the coefficient for predictor variable ’property_type', representing the change in log-odds of a violation per unit increase in 'property_type'. Formula (4) specifies that $\beta_2$ follows normal distribution with mean equals 0 and variance equals $2.5^2$.

Formula (2), (3) and (4) specifies priors for the model. This prior reflects the assumption that coefficients are centered around 0, with a standard deviation of 2.5, allowing moderate variability while discouraging extreme values.

## Model Assumptions and Limitations {#sec-model-assume}

The main, underlying assumptions and their limitations for the Bayesian model is discussed below:

Firstly, the model assumes that each observation is independent of others. This assumption suggests that violations at one property should not influence violations at another property unless explicitly modeled (e.g., through random effects).

If observations are clustered and the assumption is violated (for instance, multiple observations were conducted in the same region or by the same inspector), overestimation and underestimation of standard errors and outcomes may occur.

Secondly, the model assumes linearity in the logit. This assumption stated that the log-odds of the response variable ($\text{logit}(p_i)$) are a linear function of the predictors.

If violations change non-linearly over time, the model assumption would be violated, potentially resulting in misfit of the model.

Thirdly, the model assumes priors of $\text{Norma}(0,2.5)$ for the coefficients. They were constructed on domain knowledge, and poorly chosen priors could make the estimation outcome biased.

Also, if new property_type levels are introduced, the model may not be no longer appropriate for the new situation, as fixed effects could not be feasible to compute estimates. Random effects, then, may be more appropriate.

@@

Since the response variable is binary for each case and follows a Bernoulli distribution, it would be appropriate to use Bayesian logistic regression model to estimate it.

The predictor variable as 'date_num' is modeled as fixed effect in the model. There are various reasons for this. Firstly, As 'date_num' represents the number of days for the inspection takes place, it is a numeric variable. Modeling it as a fixed effect allows the model to estimate its direct, continuous influence on the response variable. Secondly, since it applies across all observations, regardless of property type, modeling it as a fixed effect could consider its impact consistently across the entire dataset. It also makes the coefficient of 'date_num' ($beta_1$) be easier to interpret, avoids adding complexity to the model.

The other predictor variable 'property_type' is also modeled as fixed effect. Since it is categorical variable, modeling it as fixed effect allow the model to estimate a separate coefficient for each category, relative to a reference level. This would make it easier to compare between property types, in order to understand which types of properties would be observed to have violations. Moreover, since 'property_type' has only 10 levels, it is feasible to compute it as fixed effect and conduct estimation without overloading the model.

The prior of $\mbox{Normal}(0, 2.5)$ for every coefficients is also reasonable because it reflects prior knowledge that most predictors have small-to-moderate effects on the log-odds scale. This assumption would avoid overfitting to extreme values.

The above final model decision is made after considering several alternative models and variants. One alternative is a simple linear regression model (SLR), which assumes a constant linear relationship between the predictor (violation) and the response (date_num). Although the model is extremely straightforward and easy to interpret, it is far too simple for estimating and neglect the key variable of 'property_type'. More importantly, the response variable is binary, whereas SLR assumes a continuous response variable. MLR also assumes so, and therefore is considered and rejected because of the same reason.

Another alternative considered is the Bayesian model with natural splines (e.g. $f(\text{date}_i)$). While it could model complex, non-linear time trends and is appropriate for 'date_num', the expected model for this study does not primarily focus on exploratory analysis on non-linear relationships between date_num and violations. Non-linear functions would also overfit the data if not regularized properly.

While considering alternative Bayesian logistics model, the model with 'property_type' as fixed effect was also considered. This would capture variability across property types without explicitly estimating coefficients for each type, and does not provide explicit estimates for each property type. The model would be suitable if there are many property types, but the number of types is manageable to include it in the model as fixed effect.

Therefore, after considering many alternatives, the final model of Bayesian logistics model with 'date_num' and 'property_type' as fixed effect is chosen. It could directly provides explicit comparisons between property types without overfitting the model, and meet the focus of the research question.

@@

Our results of the Bayesian logistic regression model are summarized in @tbl-modelresults. The table provides information of coefficients and main summary statistics.

The table firstly lists estimates of coefficients for all parameters. The first line suggests an intercept of -6.72 for the model, indicating very low baseline odds of a violation for the reference property type when date_num = 0. The second line presents that the estimated coefficient for date_num is 0.02, which is very small and suggests that the inspection time interval has negligible effect on the violation outcome of observations.

>>>>

In recent years, fire safety inspections in residential buildings have gained increasing attention from the public due to their role in ensuring the safety of residents and reducing the risks of fire-related accidents. Toronto Fire Services (TFS) fire prevention and enforcement strategies, which are guided by the Fire Protection and Prevention Act, 1997 (FPPA) and regulations under it (such as Ontario Fire Code), aim to enhance public safety and awareness [@fire_info].

In order to achieve this goal, TFS has been conducting inspections across a wide range of residential properties throughout the city to ensure properties' compliance with fire safety regulations. However, despite these efforts, there remains limited understanding of the factors influencing the likelihood of fire safety violations across different property types and inspection duration. As a result, targeted interventions have not been effectively implemented for high-risk buildings. Therefore, this paper aims to analyze the impact of both inspection duration and property type on the likelihood of violations during fire safety checks. By doing so, the study aims to develop a model that provides predictions of violations based on specific features of the properties and inspection characteristics, thereby improving TFS's ability to prioritize resources and focus on properties that require the most attention.

The focus of this study is to investigate how different property types and inspection duration contribute to fire safety violations in Toronto, using a Bayesian logistics regression model to conduct predictions. The predictor variables examined include the time interval of the inspection process, and the type of property being inspected. By exploring these factors, the paper aims to predict the likelihood of violations, conducting predictions of which property types are at higher risk for violations and how inspection duration might influence these outcomes. The estimand is described in detail in @sec-estimand.

As included at the end of the analysis, the model indicates that the duration of the inspection has a negligible impact on the likelihood of violations, while the variable of property type is a more significant predictor of likelihood of violations. The model predicts with high confidence that most properties would have a high probability of being observed to have a violation of fire safety regulations. This confident prediction highlights the utility of the model in identifying properties at high risk for violations, helping to conduct targeted inspections and interventions where they are most needed. Moreover, the analysis further suggests that different property types have varying impacts on the likelihood of fire safety violations. The model confidently predicts that among ten property types inspected, "High Rise," "Rooming House," and "Group Home" are more likely to have violations. As a result, these property types should receive more targeted inspections and interventions. Along with "Low Rise" and "Hotel & Motel," these property types show strong evidence of being associated with higher risks of violations. In contrast, the model indicates that "Nursing Home" has the lowest risk of violations among the ten property types inspected, with some uncertainty regarding the strength of this predicted relationship.

This research contributes to improving fire safety facilities in residential buildings in Toronto by providing empirical evidence and statistical predictions that supports the allocation of inspection resources based on both the property type and inspection duration. By highlighting which property types with how long inspection duration are most at risk for fire safety violations, the findings offer a valuable approach to prioritizing inspections and interventions. The results of the study have broader implications for fire safety policy, as they could be used to improve future regulatory practices and resource allocation strategies. By ensuring that inspection resources are concentrated on properties that present the greatest risk, this study would enhance the effectiveness of fire safety efforts, and help to reduce the risk of fire accidents more efficiently.

@@

Appendix includes three parts:[Appendix -@sec-appx-survey] describes detailed idealized sampling methodology and survey designed for residents to collect fire safety feedback; [Appendix -@sec-appx-data] provides additional data details in data cleaning process; [Appendix -@sec-model-details] presents additional details during modeling process, including posterior predictive check [Appendix -@sec-ppc], comparison between posterior and prior distributions [Appendix -@sec-pvp], and diagnostics [Appendix -@sec-diagnostics].

@@

The dataset captures data related to the results of fire safety inspections conducted by Toronto Fire Services (TFS) in residential buildings in Toronto, which is guided by the Fire Protection and Prevention Act, 1997 (FPPA) and regulations under the FPPA, like the Ontario Fire Code [@fire_info]. It includes information on properties where violations have been found by inspectors from TFS and required to correction to comply with the Ontario Fire Code at the time, as well as properties with no observable violations [@fire_data].

The data collection process was conducted by TFS Fire Inspectors. Inspectors assess tangible aspects of fire safety based on observable and measurable attributes of the physical environment within the properties. This includes evaluating factors such as proper storage of combustible materials, the functionality of fire alarms and fire pumps, the maintenance records for fire safety equipment and so on, all in accordance with the regulations under FPPA.

After the inspections were completed, the observations have been translated into records and written into collected data. The raw data provided by Open Data Toronto includes three main aspects: property identifier (such as address, property type), inspection information (such as inspection start and end dates), and inspection outcomes (pass or fail status for safety checks, and details of violations).

Note that this dataset only includes 'closed' cases, referring to those whose inspection processes have been completed. The ongoing inspections are not presented in the dataset, which may result in a potential loss of information. This limitation would be further discussed in detail in the @sec-discussion.

@@

The outcome variable for the model is the violation status of the properties. In other words, it is the variable that represents whether the properties inspected were found to be in violation of fire safety regulations by a TFS Inspector.

In the raw dataset, columns 'VIOLATION_FIRE_CODE', 'VIOLATION_DESCRIPTION' and 'VIOLATIONS_ITEM_NUMBER' indicate the outcome of the inspections. The first two columns provide detailed information about the fire code under which violation was noted, while the latter presents the order number of violations by code and shows 0 if no violations were observed.

To simplify the analysis, a new column, 'violation', is created. This column contains binary values, with 1 indicating that a violation was observed and 0 indicating no violation. Thus, the outcome variable is binary.

```{r}
#| label: fig-violation
#| fig-cap: Counts of Properties with Violations versus Those without
#| echo: false
#| warning: false
#| message: false

cleaned_data |>
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  labs(
    x = "Violation Status",
    y = "Counts",
    fill = "Violation Status" # Legend title
  ) +
  scale_fill_manual(
    values = c("skyblue", "orange"),
    labels = c("No Violation Observed (Denoted as 0)", "Violation Observed (Denoted as 1)")
    ) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

@fig-violation displays the visualized counts of properties with violations versus those without. The height of each bar shows the total number of cases corresponding to each category. The orange bar is significantly taller, indicating that among all cases selected for analysis, violations were reported more frequently in the dataset (9,829 cases out of 13,193). In contrast, the blue bar is shorter, reflecting fewer cases without violations (3,364 cases out of 13,193).

The data contains a higher number of cases with violations (1) compared to those where no violations were observed (0), suggesting that violations are relatively common.

@@

### Inspection Duration

One predictor included in the model is the inspection duration. It refers to the time interval between the start and end of the fire inspections conducted by the TFS inspectors for each case.

In the raw dataset, there are two columns recording the dates: 'INSPECTIONS_OPENDATE' and 'INSPECTIONS_CLOSEDDATE'. The former represents the date TFS opened the inspection file, while the latter denotes the date TFS closed inspection file or the date enforcement proceedings ended (whichever is later).

During the data cleaning process, these dates are converted into time interval to enable easier analysis. A new column, 'date_num', was created by calculating the time difference between the end and start times of the inspection, in the unit of days.

```{r}
#| label: fig-date
#| fig-cap: Distribution of Inspection Duration
#| echo: false
#| warning: false
#| message: false

cleaned_data |> 
  ggplot(aes(x = date_num)) + 
  geom_histogram(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(x = "Inspection Duration (Days)",
       y = "Counts") +
  theme_minimal()
```

@fig-date illustrates the distribution of inspection duration. The X-axis represents the number of days for each inspection case, while the Y-axis represents the counts. The distribution peaks at low intervals, with the highest bar near 0 days, indicating that a large number of properties have very short inspection time intervals. The right-skewed shape of the distribution suggests that as the inspection duration increases, the counts of cases decrease stably. The long tail on the right side of the distribution means that there also exist cases with very long duration in the dataset, extending up to 1,500 days or more.

@@

The second predictor variable in the model is property type. It refers to the occupancy type of the property where the inspection took place. This variable is directly included in the raw dataset.

To visualize the distribution of property types, two bar plots are created separately in @fig-type. The left panel presents the number of occurrences of "High Rise" and "Low Rise" properties, with the bar representing "High Rise" relatively taller, indicating a higher count compared to "Low Rise." This suggests that the inspection results and data collection focus more heavily on high-rise buildings.

The right panel displays the other categories of properties, including "Detention," "Group Home," "Group Home (VO)," "Hospital," "Hotel & Motel," "Nursing Home," "Residential Care," and "Rooming House." The "Rooming House" category has the highest count among these property types, while categories of "Detention" and "Group Home" have relatively lower counts, suggesting that they are less represented in the dataset.

In general, the distribution of the property types is diverse, with some types being far less frequent than others. High-rise properties dominate the dataset, while rooming houses are the most frequent within other categories. This pattern may indicate an emphasis on inspecting high-rise buildings and rooming houses than those of other property types.

```{r}
#| label: fig-type
#| fig-cap: Distribution of Property Types
#| echo: false
#| warning: false
#| message: false
#| fig-subcap: ["High Rise versus Low Rise", "Other Types"]
#| layout-ncol: 2

@@

# Create bar plot for high rise properties versus low rise properties
data_rise |>
  ggplot(
    aes(x = as.factor(property_type), 
        fill = as.factor(property_type))
    ) +
  geom_bar(color = "black") +
  labs(
    x = "Property Type",
    y = "Counts",
    fill = "Property Type"
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

# Create bar plot for other property types
data_other |>
  ggplot(
    aes(x = as.factor(property_type), 
        fill = as.factor(property_type))
    ) +
  geom_bar(color = "black") +
  labs(
    x = "Property Type",
    y = "Counts",
    fill = "Property Type" 
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
    axis.text.y = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

@@

@fig-violationbytype displays faceted bar plots illustrating the distribution of violation statuses (0 for no violation, 1 for violation) across different property types. Each subplot represents a different property type.

In general, the property types of "Group Home," "High Rise," "Low Rise," and "Rooming House" show a higher proportion of violations (1). The type "Hotel & Motel" has relatively balanced counts between 0 and 1, but violations (1) are slightly more frequent. These property types may require more attention or stricter inspections.

On the other hand, other property types, especially "Detention" and "Nursing Home," show a much lower proportion of violations, suggesting a lower relative violation rate.

@@

cleaned_data |> 
  ggplot(aes(x = as.factor(violation), fill = as.factor(violation))) +
  geom_bar(color = "black") +
  facet_wrap(~ property_type, scales = "free_y") + # Create one subplot per property type
  labs(
    x = "Violation Status",
    y = "Counts",
    fill = "Violation Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_blank(), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 8), 
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    panel.spacing = unit(1, "lines")
  ) +
  scale_fill_manual(values = c("skyblue", "orange"))
```

@@

Firstly and most importantly, from a statistical perspective, the goal is to estimate the detection of violations during an inspection based on the time interval and property type.

Secondly, from a more practical standpoint, by predicting which property types and inspection duration are more likely to result in violations, a more efficient inspection process could be developed. Identifying certain property types that have high risk to violate the fire safety regulations could provide guidelines for interventions and support implementation of targeted and focused actions.

The model presented in the paper is Bayesian logistics regression model designed to predict the violation status (denoted as 'violation') based on the fixed effects of inspection duration (denoted as 'date') and property type (denoted as 'type'). It is designed to forecast the inspection results based on these two predictor variables.

The following sections define, explain and justify the model and its variables. They also discuss underlying assumptions, potential limitations, the software used for implementation, and provide evidence of model validation and checking.

@@

The Bayesian model was implemented using the R programming language [@citeR], specifically utilizing the `rstanarm` package by @citerstanarm. This package provides an interface for fitting Bayesian regression models using Stan, and the models are fit using its functionality.

Define $y_i$ as the binary response variable for observation $i$. $y_i=1$ indicates that a violation has been observed, while $y_i=0$ indicates that no violation occurs. This follows a Bernoulli distribution:

@@

Then define the inspection time interval for observation $i$ as 'date' and the property type as 'type'.

The Bayesian model could be expressed with the following mathematical formulations:

@@

- $\text{logit}(p_i) = \log\left(\frac{p_i}{1 - p_i}\right)$ represents the log-odds of $p_i$.
- $\beta_0$ is the intercept term of the Bayesian logistic regression, representing the baseline log-odds of a violation when $date = 0$ and 'type' is the reference category. Formula (2) specifies that $\beta_0$ follows normal distribution with a mean of 0 and a variance of $2.5^2$.
- $\beta_1$ is the coefficient for predictor variable 'date', representing the change in log-odds of a violation for per unit increase in 'date', i.e., the inspection time interval. Formula (3) specifies that $\beta_1$ follows normal distribution with a mean of 0 and a variance of $2.5^2$.
- $\beta_2$ is the coefficient for the predictor variable 'type', representing the change in log-odds of a violation for each unit increase in 'type'. Formula (4) specifies that $\beta_2$ follows normal distribution with a mean of 0 and a variance of $2.5^2$.

Formula (2), (3) and (4) specifies priors for the model. These priors reflect the assumption that the coefficients are centered around 0, with a standard deviation of 2.5, allowing for moderate variability while discouraging extreme values.

## Model Assumptions {#sec-model-assume}

The main assumptions for the Bayesian model and situations where the model may not be appropriate are discussed below. Limitations of the model are covered in detail in @sec-limit.

- Independence of Observations

Firstly, the model assumes that each observation is independent of others. This assumption suggests that violations at one property should not influence violations at another property, unless explicitly modeled (e.g., through random effects).

If observations are clustered (e.g., multiple observations in the same region or by the same inspector were recorded in the dataset), this assumption may be violated, leading to potential overestimation or underestimation of standard errors and outcomes.

- Linearity in the Logit

Secondly, the model assumes linearity in the logit. This assumption states that the log-odds of the response variable ($\text{logit}(p_i)$) are a linear function of the predictors.

If violations change in a non-linear manner over time, this model assumption would be violated, potentially resulting in misfit of the model.

- Prior Distributions

Thirdly, the model assumes priors of $\text{Norma}(0,2.5)$ for the coefficients, based on domain knowledge. Poorly chosen priors could introduce bias into the estimation results.

Also, if new levels of property type ('type') levels are introduced, the model may be no longer suitable for the new situation, as fixed effects may not be feasible for estimating the new categories. In such cases, random effects might be a more appropriate alternative.

@@

Since the response variable is binary for each case and follows a Bernoulli distribution, a Bayesian logistic regression model is appropriate for estimating it.

The predictor variable of inspection duration ('date') is modeled as a fixed effect in the model. There are several reasons for this choice. Firstly, 'date' represents the number of days on which the inspection processed, making it a numeric variable. Modeling it as a fixed effect allows the model to estimate its direct, continuous influence on the response variable. Secondly, since 'date' applies across all observations regardless of property type, modeling it as a fixed effect allows its impact to be considered consistently across the entire dataset. This also makes the coefficient of 'date' ($beta_1$) be easier to interpret and avoids adding unnecessary complexity to the model.

The other predictor variable, property type ('type'), is also modeled as a fixed effect. Since it is categorical variable, modeling it as a fixed effect enables the model to estimate a separate coefficient for each category, relative to the reference level. This makes it easier to compare different property types, in order to understand which types of properties are more likely to have violations. Moreover, since 'property_type' has only 10 levels, it is feasible to compute it as a fixed effect and develop estimation without overloading the model.

The prior of $\mbox{Normal}(0, 2.5)$ for each coefficient is reasonable because it reflects prior knowledge that most predictors have small-to-moderate effects on the log-odds scale. This assumption helps prevent overfitting by discouraging extreme values.

The final model choice was made after considering several alternative models. One alternative is a simple linear regression (SLR) model, which assumes a constant linear relationship between the predictor (violation) and the response (date). While straightforward and easy to interpret, this model is too simplistic for estimation, neglecting the important variable of property type (type). More importantly, the response variable is binary, and SLR assumes a continuous response, making it unsuitable for this case. Multiple linear regression (MLR) also assumes a continuous response variable, so it was similarly rejected.

Another alternative considered was a Bayesian model with natural splines (e.g. $f(\text{date}_i)$), which could model complex, non-linear time trends and would be appropriate for 'date'. However, the expected model for this study does not primarily focus on exploratory analysis on non-linear relationships between 'date' and 'violation'. Non-linear functions would also overfit the data if not regularized properly.

Additionally, a Bayesian logistic model with property type ('type') as a random effect was considered. This would capture variability across property types without explicitly estimating coefficients for each type. While this model would be suitable if there were many property types, it was unnecessary given the manageable number of property types in this case.

Therefore, after considering these alternatives, the final model of Bayesian logistics model with inspection duration ('date') and property type ('type') as fixed effects was chosen. This model allows for direct, explicit comparisons between property types without overfitting the model, and aligns with the focus of the research question.

@@

The results of the Bayesian logistic regression model are summarized in @tbl-modelresults. The table provides information about the coefficients and main summary statistics.

The table firstly lists estimates of coefficients for all parameters. The first line suggests an intercept of -6.72 for the model, indicating very low baseline odds of a violation for the reference property type when $date = 0$. The second line shows that the estimated coefficient for predictor 'date' is 0.02, which is very small and suggests that the inspection duration has a negligible effect on the violation outcome of observations.

The table then presents the coefficients for every property type respectively. All property types show large positive coefficients, meaning that they increase the log-odds of violations and have a higher risk of violation compared to the reference property type. For example, for the property type of "Group Home", the log-odds of violation increases by 5.88; for the property type of "High Rise", the increase is 6.21; for the property type of "Hospital", the increase is 4.67; and so on. It could be observed that "Nursing Home" has the smallest coefficient, suggesting that it has the smallest effect on the likelihood of violations. However, it still contributes positively to the log-odds of a violation.
